---
title: "i4TB Didadic session: Plotting with ggplot"
output: html_notebook
---

This notebook will cover plotting techniques using the R library `ggplot2`. We will prioritize plots that are hard to generate in excel. The code provided here is designed to facilitate customization for new datasets.

For more information on the tools mentioned, please consult these references:

#### General references ####
[ggplot homepage](https://ggplot2.tidyverse.org/)

[ggplot cheatsheet](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

[ggplot essentials](http://www.sthda.com/english/wiki/ggplot2-essentials)

[R cookbook reference](http://www.cookbook-r.com/Graphs/)

[R Graph Gallery](https://www.r-graph-gallery.com/ggplot2-package.html)

#### References on modifications ####
[Shapes](http://sape.inf.usi.ch/quick-reference/ggplot2/shape)

[Picking colors](https://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

[Using math symbols](https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html)

[More on using math symbols](http://csrgxtu.github.io/2015/03/20/Writing-Mathematic-Fomulars-in-Markdown/)




# 1. Installing and loading packages
First, we will need to load the necessary packages for this session. If there are any packages that need to be installed, use the install scripts once. *Note:* If you already have these installed, no need to run this again.
```{r}
# Code for installation only. Execute this only once.
install.packages('ggplot2')
install.packages('ggExtra')
install.packages('RColorBrewer')
install.packages('Rmisc')
install.packages('tidyverse')

# Loading the packages
library(ggplot2)
library(ggExtra)
library(RColorBrewer)
library(Rmisc)
library(tidyverse)
```

# 2. Loading a dataset

Next, we want to input our datasets. We're importing the data as a `.csv` file (which can be generated from Excel). 

```{r}
# Importing from CSV
data = read_csv('DataToPlot1.csv')

View(data)
```

For the purposes of this first set of plot, we'll want to focus on an individual gene: TBCD.

```{r}
# Filtering by the gene TBCD
DataToPlot = filter(data, GeneID == "TBCD")
View(DataToPlot)
```

The data frame `DataToPlot` is in a format that is compatible with `ggplot`. 
# 3. Box plot


Here's the basic command to generate a box plot from the data in `DataToPlot`.

```{r}
# Basic Box Plot
ggplot(data = DataToPlot, aes(x=TB, y=Expression)) + 
  geom_boxplot() # Setting plot type to be boxplot
```
We can add all sorts of customizations to this plot to make it look prettier:

```{r}
# Box Plot with Aesthetic Customizations
ggplot(data = DataToPlot, aes(x = TB, y=Expression, fill = TB)) + 
  geom_boxplot() + # Setting plot type to be boxplot
  xlab('Infection Type') + # x label
  ylab('Expression [counts]') + # y label
  ggtitle('TBCD Expression') + # title
  scale_fill_brewer(name="Infection\nType", palette = 'Set1') + # setting different color schemes and customizing legend title
  theme_bw() + # changing background color 
  theme(plot.title=element_text(hjust=0.5)) # adjust title position
```

# 4. Layering multiple graph types

You can layer multiple graph types onto the same plot

```{r}
# Box Plot + Jitter Plot
ggplot(data = DataToPlot, aes(x=TB, y=Expression)) + 
  geom_boxplot(outlier.shape = NA, outlier.size = 0) + # Setting plot type to be boxplot, and removing outlier points
  geom_jitter(height=0, width=0.2, aes(color=TB)) + # Layering on a jitter plot of the individual replicates
  scale_color_brewer(name="Infection\nType", palette = 'Dark2') + # setting different color schemes and customizing legend title
  xlab('Infection Type') + # x label
  theme_bw() # changing background color
```

# 5. Faceting

If you want to generate a group of related plots, each of which plotting a subset of the data (based on a specific grouping), you can do this easily using faceting.

In this example, we're generating a family of faceted plots, each of which plots the expression of for a different patient.

```{r}
# Jitter Plot with Faceting
ggplot(data = DataToPlot, aes(x=TB, y=Expression)) + 
  geom_jitter(aes(color = TB)) + # Setting plot type to be boxplot
  facet_grid(. ~ ptID,labeller = label_both) + # creating facets
  scale_color_discrete(name="Infection\nType") + # customizing legend title
  ggtitle('TBCD Expression') + # title
  theme_bw() + # changing background color
  theme(plot.title=element_text(hjust=0.5)) # adjust title position
```

# 6. Plotting distributions

Distributions can be represented as histograms and density plots:

```{r}
# Basic Histograms and Density Plots
ggplot(data = DataToPlot, aes(x=Expression)) + 
  geom_histogram(binwidth=.2) + # Setting plot type to be historgram
  geom_density() + # Setting plot type to be density plot
  xlab('TBCD Expression') +
  ylab('Frequency')
```

Can also plot by group:

```{r}
# Histograms colored by group
ggplot(data = DataToPlot, aes(x=Expression,fill = TB)) + 
  geom_histogram(binwidth=.2, position="dodge") + # Setting plot type to be histogram with interleaved bars
  xlab('TBCD Expression') +
  ylab('Frequency')
```

```{r}
# Density plots colored by group
ggplot(data = DataToPlot, aes(x=Expression,fill = TB)) + 
  geom_density(alpha=.5) + # Setting plot type to be density plot
  xlab('TBCD Expression') +
  ylab('Frequency')
```


# 7. Bonus: Adding marginal histograms to scatter plots.

You can also plot disributions in 2 dimensions. 

To demonstrate this, we'll first construct a different data frame, which has the expression values of TBCD and NOC2L in 2 separate columns.

The `filter` function extracts only the rows with the genes of interest, and the `pivot_wider` function reshapes the data frame to have expression of the two genes represented in separate columns.

```{r}
# Creating a data frame of sample gene expression with gene TBCD and NOC2L represented in separate columns.
DataToPlot2 = pivot_wider(data, names_from = GeneID, values_from = Expression) 
View(DataToPlot2)
```


Now to generate an x/y scatter plot of these two genes. Importantly, you need to save the `ggplot` function output to a variable.

```{r}
# Scatter Plot
p = ggplot(data = DataToPlot2, aes(x=TBCD, y=NOC2L)) + 
  geom_point(aes(color = TB)) + # Setting plot type to be scatter plot
  scale_color_discrete(name="Infection\nType") # setting legend label
p # Showing the plot
```

We can add to this marginal plot visualizations of the distributions of each individual gene to the scatter plot:

```{r}
# Adding a marginal plot
p1 = ggMarginal(p, type="density", groupColour = TRUE, groupFill = TRUE) # adding marginal density plots
p1 # showing the plot
```

